SOURCE SEPARATION METHODS FOR COMPUTER-ASSISTED ORCHESTRATION

### 2021-06-03 source_separation/paper/paper.tex
Move discussion of dynamic orch to future work

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}
%
\begin{keywords}
computer-assisted orchestration, sound source separation, Orchidea
\end{keywords}
%
\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes. In particular, target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints~\cite{Maresz2003}.

  Computer-assisted orchestrations can be divided into two categories: static and dynamic. Static orchestration considers the target as a single vector of timbral features and creates a single onset of note(s) as the solution. Dynamic orchestration considers a time series of features, which allows it to create an orchestration that evolves over time with multiple onsets. It uses time segmentation to cut the target in time into multiple segments, and orchestrates each segment.

  Orchidea is a framework and set of tools that is currently considered the state-of-the-art for computer-assisted orchestration. It embeds the target in a feature space and uses a jointly-optimized heuristic and constraint solver to find a combination of samples that best match the target~\cite{Cella18, Cella2020}. Orchidea is able to do both static and dynamic orchestration. In order to better observe the effects of source separation, we perform only static orchestrations in this paper. The reasoning for this is that we wish to disentangle the two difficult problems of source separation and time segmentation. By only performing static orchestrations, we reduce the complexity of the problem by removing time segmentation and can better observe the effects of source separation.

@@

  For example, consider a target that has a continuous drone sound throughout and a melody playing above this drone. Without source separation, Orchidea would detect each note of the melody as a new onset and cut the target in time at each new note. However, this segmentation in time would also effect the drone, cutting it in time and forcing it to have a new onset each time the melody changes notes. The resulting orchestration would have multiple onsets for the drone, even though only one onset is needed at the beginning of the drone. When source separation is applied to this target, the drone and melody would be orchestrated separately. Therefore, the drone could be orchestrated by a single onset that lasts for the duration of the drone. At the same time, the melody could have as many new onsets as needed without affecting the drone. The orchestration would be greatly improved by removing the unnecessary onsets in the orchestration of the drone. Although this example is a dynamic orchestration, and we perform only static orchestrations, this paper represents a first step toward achieving dynamic orchestration with source separation.

@@

    As we stated in Section \ref{sec:introduction}, we chose static orchestration instead of dynamic orchestration to avoid entangling the problems of time segmentation and source separation. Future work could include applying source separation to dynamic targets and performing dynamic orchestration.

>>>>

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}
%
\begin{keywords}
computer-assisted orchestration, sound source separation, Orchidea
\end{keywords}
%
\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes. In particular, target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints~\cite{Maresz2003}. Static orchestration is a type of orchestration that considers the target as a single vector of timbral features and creates only a single onset of note(s) as the solution, no matter if the target contains multiple onsets of sounds. In our experiments, we only perform static orchestrations.

  Orchidea is a framework and set of tools that is currently considered the state-of-the-art for computer-assisted orchestration. It embeds the target in a feature space and uses a jointly-optimized heuristic and constraint solver to find a combination of samples that best match the target~\cite{Cella18, Cella2020}. All of our orchestrations are done using Orchidea's command line tools for batch processing, which is available for download at \url{http://www.orch-idea.org}

@@

  For example, consider a target that has a continuous drone sound throughout and a melody playing above this drone. Without source separation, Orchidea would detect each note of the melody as a new onset and cut the target in time at each new note. However, this segmentation in time would also effect the drone, cutting it in time and forcing it to have a new onset each time the melody changes notes. The resulting orchestration would have multiple onsets for the drone, even though only one onset is needed at the beginning of the drone. When source separation is applied to this target, the drone and melody would be orchestrated separately. Therefore, the drone could be orchestrated by a single onset that lasts for the duration of the drone. At the same time, the melody could have as many new onsets as needed without affecting the drone. The orchestration would be greatly improved by removing the unnecessary onsets in the orchestration of the drone.

@@

    As we stated in Section \ref{sec:introduction}, we performed only static orchestration in our experiments. In contrast to static orchestration, dynamic orchestration considers a time series of features, which allow the orchestration to evolve over time and contain multiple onsets of notes. Orchidea is able to perform dynamic orchestration by cutting the target in time into multiple segments, and then orchestrating each segment. However we chose to use only static orchestrations in order to better observe the effects of source separation. The reasoning for this is that we wish to disentangle the two difficult problems of source separation and time segmentation. By only performing static orchestrations, we reduce the complexity of the problem by removing time segmentation and can better observe the effects of source separation. Future work could include applying source separation to dynamic targets and performing dynamic orchestration.
